Carryover of a saddle-node bifurcation after transformation of a parameter into a variable: Appendix notes

### 2020-08-07 Manuscript/notes.tex
notes

\title{Carryover of a saddle-node bifurcation after transformation of a parameter into a variable}
\author{Carlos Contreras, Gustavo Carrero, Gerda de Vries}
% \address{University of Alberta}

@@

\begin{abstract}
To do.

Saddle-Node Bifurcation, Carryover, Bifurcations
\end{abstract}

\linenumbers

@@

% Saddle-node bifurcations are very common in biology, medicine and ecology [references, eg. Li et al. 2019], explaining hysteresis and critical transitions.

In the general case, we are interested in the conditions such that a saddle-node bifurcation for the system
\begin{equation}
    \dot z = f(z;\mu_{1},\mu_{2}),
    \label{equ:OriginalSystem}
\end{equation}
where $z\in\mathbb{R}^{n}$ and $\mu_{1},\mu_{2}\in\mathbb{R}$, manifests itself after we extend the system by transforming one of the parameters, for example, $\mu_{1}$, into a variable. This defines the extended system
\begin{equation}
    \left. \begin{aligned}
    \dot z &= f(z,\mu_{1}; \mu_{2}), \\
    \dot \mu_{1} &= g(\mu_{1}; \mu_{2}),
    \end{aligned} \right.
    \label{equ:ExtendedSystem}
\end{equation}
where a saddle-node bifurcation occurs and $g(\mu_{1}; \mu_{2})$ is the vector field of the new variable $\mu_{1}$. We call this saddle-node the carryover of a saddle-node bifurcation in the original system \eqref{equ:OriginalSystem} to the extended system \eqref{equ:ExtendedSystem}. Note that the function $g$ does not depend on $z$. The case where $g$ depends on $z$ is beyond the scope of this chapter, but we briefly discuss this case in Section~\ref{sec:Discussion}.

Saddle-node bifurcations in $\mathbb{R}^{n}$ are characterized by three conditions: singularity, nondegeneracy, and transversality conditions. They guarantee the creation (or destruction) of two equilibria as one parameter crosses the bifurcation value. This is summarized in the following results taken from \citet[Ch. 8]{Meiss2007}.

\begin{theorem}[saddle node]
    \label{the:SaddleNodeBifurcationND}
    Let $f\in C^2(\mathbb{R}^n\times \mathbb{R}^k,\mathbb{R}^n)$, and suppose that $f(z;\mu$) satisfies
    \begin{equation}
        f(0;0)=0, \quad \mathrm{spec}(D_z f(0;0))=\{ 0, \lambda_2, \lambda_3, \ldots, \lambda_n : \lambda_k \neq 0, k\neq 1 \}.
        \label{equ:SaddleNodeBifurcationND:Singularity}
    \end{equation}
    Choose coordinates so that $D_zf(0;0)$ is diagonal in the zero eigenvalue and set $z=(x,y)$ where $x\in \mathbb{R}^1$ corresponds to the zero eigenvalue and $y\in \mathbb{R}^{n-1}$ are the remaining coordinates. Then 
    \begin{equation}
        \begin{split}
            \dot x &= f_1(x,y;\mu), \\
            \dot y &= My + f_2(x,y;\mu),
        \end{split}
        \label{equ:SaddleNodeBifurcationND:System}
    \end{equation}
    where $f_{1}(0,0;0)=0$, $f_{2}(0,0;0)=0$, $D_z f_{1}(0,0;0)=0$, $D_z f_{2}(0,0;0)=0$, and $M$ is an invertible matrix. Suppose that 
    \begin{equation}
        D_{xx}f_1(0,0;0)=c\neq 0.
        \label{equ:SaddleNodeBifurcationND:Nondegeneracy}
    \end{equation}
    Then there exists an interval $I(\mu)$ containing $0$, functions $y=\eta(x;\mu)$ and extremal value $m(\mu)=\mathop{\mathrm{Ext}}_{x\in I(\mu)}[f_1(x;\eta(\mu);\mu)]$, and a neighborhood of $\mu=0$ such that if $m(\mu)c>0$ there are no equilibria and if $m(\mu)c<0$ there are two. Suppose that $M$ has a $u$-dimensional unstable space and an $(n-u-1)$-dimensional stable space. Then, when there are two equilibria, one has a $u$-dimensional unstable manifold and an $(n-u)$-dimensional stable manifold and the other has a $(u+1)$-dimensional unstable manifold and an $(n-u-1)$-dimensional stable manifold.
\end{theorem}

Equations \eqref{equ:SaddleNodeBifurcationND:Singularity} and \eqref{equ:SaddleNodeBifurcationND:Nondegeneracy} are the singularity and nondegeneracy conditions, respectively. They are necessary conditions for the function $f_{1}$ to be zero up to the zero- and first-order approximations about the bifurcation point, but nonzero in the second-order approximation. The function $y=\eta(x;\mu)$ allows us to reduce the dynamics in a neighborhood of the bifurcation point to one-dimension, i.e.,
\begin{equation*}
    \dot x = f_{1}(x,\eta(x;\mu);\mu).
\end{equation*}
The extremal value function $m(\mu)$ determines a single condition on the parameters, $m(\mu)=0$, along which two equilibria are created (or destroyed). Having one condition on the parameters means that the bifurcation that takes place has codimension-one. In order to be a saddle-node bifurcation, the equilibria need to be created as a some combination of the parameters crosses the bifurcation point. This can be guaranteed with a simple condition.

\begin{corollary}
    \label{the:SaddleNodeBifurcationNDCorollary}
    If $\mu_1$ is a single parameter such that 
    \begin{equation}
        D_{\mu_1}f_1(0,0;0)\neq 0,
        \label{equ:SaddleNodeBifurcationND:Transversality}
    \end{equation}
    then a saddle-node bifurcation takes place when $\mu_1$ crosses zero.
\end{corollary}

Equation \eqref{equ:SaddleNodeBifurcationND:Transversality} is the transversality condition that guarantees that $m(\mu)=0$ is crossed transversally as $\mu_{1}$ crosses zero. Note that $\mu_{1}$ is an arbitrary parameter, and that the transversality condition can hold for several parameters at the same time. We only consider saddle-node bifurcations that take place as a single parameter crosses the bifurcation point.

In the context of this chapter, we only consider two parameters for the sake of simplicity, i.e., $\mu=(\mu_{1},\mu_{2})\in\mathbb{R}^{2}$. We show that as long as a saddle-node bifurcation takes place in system \eqref{equ:OriginalSystem} for at least one of the parameters, the extended system \eqref{equ:ExtendedSystem} also has a saddle-node bifurcation for the other parameter (the extended variable does not have to be a bifurcation parameter in the original system) under some singularity and transversality conditions.

The Implicit Function Theorem is an essential tool in this chapter and in the study of saddle-node bifurcations in general. For example, the function $y=\eta(x;\mu)$ in Theorem~\ref{the:SaddleNodeBifurcationND} is consequence of this theorem. The following form of the Implicit Function Theorem is taken from \citet[Ch. 8]{Meiss2007}.
\begin{theorem}[implicit function]
    \label{the:ImplicitFunction}
    Let $U$ be an open set in $\mathbb{R}^{n}\times\mathbb{R}^{k}$ and $F\in C^{r}(U,\mathbb{R}^{n})$ with $r\geq 1$. Suppose there is a point $(x_{0},\mu_{0})\in U$ such that $F(x_{0};\mu_{0})=c$ and $D_{x}F(x_{0};\mu_{0})$ is a nonsingular matrix. Then there are open sets $V\subset \mathbb{R}^{n}$ and $W\subset \mathbb{R}^{k}$ and a unique $C^{r}$ function $\xi(\mu):W\mapsto V$ for which $x_{0}=\xi(\mu_{0})$ and $F(\xi(\mu);\mu)=c$.
\end{theorem}

This chapter is structured as follows. In Section \ref{sec:1DCase}, we look at the case where system \eqref{equ:OriginalSystem} is one dimensional, in which case we replace $z\in\mathbb{R}^{n}$ with $x\in\mathbb{R}^{1}$. We find conditions such that the two-dimensional extended system \eqref{equ:ExtendedSystem} has a saddle-node bifurcation. Moreover, we provide a graphical way to easily verify such conditions: the nullcline of the new equation ($g=0$ in the extended system \eqref{equ:ExtendedSystem}) has to intersect transversally the two-parameter bifurcation curve of the original system \eqref{equ:OriginalSystem}. Then, we apply our results to a few illustrative examples. In Section~\ref{sec:NDCase}, we expand our results to the $n$-dimensional case and apply our results to one illustrative example. Finally, in Section~\ref{sec:Discussion}, we discuss our results and further research.

\section{One-dimensional case}
\label{sec:1DCase}

In this section, we focus on the case where the variable $z$ in the system \eqref{equ:OriginalSystem} is one-dimensional, i.e., the system
\begin{equation}
    \dot x = f(x;\mu_{1},\mu_{2}),
    \label{equ:1DCase:OriginalSystem}
\end{equation}
where $x\in\mathbb{R}$, $\mu_{1},\mu_{2}\in\mathbb{R}$, and $f$ is a sufficiently smooth function on $(x,\mu_{1},\mu_{2})$. This system is extended by transforming one of the parameters, for example, $\mu_{1}$, into a variable to define the extended system
\begin{equation}
    \left. \begin{aligned}
    \dot x &= f(x,\mu_{1}; \mu_{2}), \\
    \dot \mu_{1} &= g(\mu_{1}; \mu_{2}),
    \end{aligned}\right.
    \label{equ:1DCase:ExtendedSystem}
\end{equation}
where $g(\mu_{1}; \mu_{2})$ is the sufficiently smooth vector field of the new variable $\mu_{1}$. We want to find conditions for the carryover of a saddle-node bifurcation in the original \eqref{equ:1DCase:OriginalSystem} to the extended system \eqref{equ:1DCase:ExtendedSystem}.

Suppose, without loss of generality, that the saddle-node bifurcation for \eqref{equ:1DCase:OriginalSystem} occurs at the origin as one parameter, $\mu_{1}$, crosses zero. That is, $f$ at $(x,\mu_{1}, \mu_{2})=(0,0,0)$ satisfies the singularity conditions
\begin{equation}
    \left\{ \begin{aligned}
    &f(x;\mu_{1},\mu_{2})=0, \\
    &D_{x}f(x;\mu_{1,\mu_{2}})=0,
    \end{aligned}\right.
    \label{equ:1DCase:SingularitySystem}
\end{equation}
and the nondegeneracy and transversality conditions
\begin{equation}
    \left\{ \begin{aligned}
    &D_{xx}f(x;\mu_{1},\mu_{2})\neq 0, \\
    &D_{\mu_{1}}f(x;\mu_{1},\mu_{2})\neq 0.
    \end{aligned}\right.
    \label{equ:1DCase:NondegeneracyAndTransversalitySystem}
\end{equation}

Note that system \eqref{equ:1DCase:SingularitySystem} has two equations in $\mathbb{R}^{3}$ with coordinates $(x,\mu_{1}, \mu_{2})$ and Jacobian
\begin{equation*}
    J = \begin{pmatrix}
         D_{x}f & D_{\mu_{1}}f & D_{\mu_{2}}f \\
         D_{xx}f & D_{x\mu_{1}}f & D_{x\mu_{2}}f
        \end{pmatrix} =
        \begin{pmatrix}
         0 & D_{\mu_{1}}f & D_{\mu_{2}}f \\
         D_{xx}f & D_{x\mu_{1}}f & D_{x\mu_{2}}f
        \end{pmatrix}.
\end{equation*}
This matrix has full rank since
\begin{equation*}
    \det \begin{pmatrix}
         D_{x}f & D_{\mu_{1}}f\\
         D_{xx}f & D_{x\mu_{i}}f
    \end{pmatrix} = 
    \det \begin{pmatrix}
         0 & D_{\mu_{1}}f\\
         D_{xx}f & D_{x\mu_{1}}f
    \end{pmatrix} = -D_{\mu_{1}}fD_{xx}f\neq 0,
%     \label{equ:1DCase:detJ}
\end{equation*}
by conditions \eqref{equ:1DCase:SingularitySystem} and \eqref{equ:1DCase:NondegeneracyAndTransversalitySystem}. The Implicit Function Theorems~\ref{the:ImplicitFunction} guarantees the existence of an interval $I$ and unique functions
\begin{equation}
    \begin{aligned}
    x &= \mathcal{X}(\mu_{2}), \\
    \mu_{1} &= \mathcal{M}(\mu_{2}), \\
    \end{aligned}
    \label{equ:1DCase:XandMfunctions}
\end{equation}
for $\mu_{2}\in I$, such that
\begin{equation*}
    \mathcal{X}(0)=0,\quad \mathcal{M}(0)=0,
\end{equation*}
and the singularity condition \eqref{equ:1DCase:SingularitySystem} is satisfied. This defines a smooth one-dimensional curve $\Gamma$ that follows the bifurcation point $(x,\mu_{1},\mu_{2})=(0,0,0)$ and is parameterized by $\mu_{2}$, i.e.,
\begin{equation}
    \Gamma = \{ (x,\mu_{1},\mu_{2}) \colon 
    x = \mathcal{X}(\mu_{2}), \mu_{1} = \mathcal{M}(\mu_{2}), \mu_{2}\in I \},
%     \mathcal{X}(0)=0, \mathcal{M}(0)=0
%     \label{equ:1DCase:Gamma}
\end{equation}
and points in $\Gamma$ satisfy \eqref{equ:1DCase:SingularitySystem}.

By continuity, we can start from $(x,\mu_{1},\mu_{2})=(0,0,0)$ and follow the points that satisfy the singularity conditions \eqref{equ:1DCase:SingularitySystem} and the nondegeneracy and transversality conditions \eqref{equ:1DCase:NondegeneracyAndTransversalitySystem} to extend $\Gamma$. If the transversality condition is violated at some point, $D_{\mu_{1}}f=0$, but the same condition is satisfied for the other parameter, $D_{\mu_{2}}f\neq0$, we apply similar arguments to parameterize $\Gamma$ by $\mu_{1}$ in that section. Hence, we can extend $\Gamma$ from the bifurcation point $(x,\mu_{1},\mu_{2})=(0,0,0)$ beyond the interval $I$ as long as the transversality condition is satisfied for at least one of the parameters (see Figure~\ref{fig:G2ModuleMass:BifurcationCurveGamma}). The projection of the extended $\Gamma$ onto the ($\mu_{1},\mu_{2}$)-plane, given by $\pi:(x,\mu_{1},\mu_{2})\mapsto(\mu_{1},\mu_{2})$, defines an implicit function
\[h(\mu_{1},\mu_{2})=0, \]
known as \textit{bifurcation boundary}, commonly found numerically using continuation (for example, the curve in Figure~(Figure in Chapter 2)c that follows the SNIC$_{Mass}$ bifurcation). Note that, although $\Gamma$ is a smooth curve, $h(\mu_{1},\mu_{2})=0$ is not necessarily smooth at every point. For more details on two-parameter bifurcations see \citet{Kuznetsov2004}.

\begin{figure}
    \begin{center}
    \includegraphics[width=0.50\linewidth]{figures/BifurcationCurveGamma.pdf}
    \end{center}
    \caption{A bifurcation curve $\Gamma$ and its corresponding bifurcation boundary $h(\mu_{1},\mu_{2})=0$ (projection onto the $(\mu_{1},\mu_{2})$-plane). Modified from Figure~8.1 in \citet{Kuznetsov2004}.}
    \label{fig:G2ModuleMass:BifurcationCurveGamma}
\end{figure}

If the extended system \eqref{equ:1DCase:ExtendedSystem} has a saddle-node bifurcation that is the carryover of the saddle-node bifurcation of interest in original system \eqref{equ:1DCase:OriginalSystem}, then this bifurcation must take place on $\Gamma$ as it is the set of points satisfying the conditions for a saddle-node bifurcation.

\begin{proposition}
    \label{the:1DCase:SaddleNodeBifurcationExtension1D}
    Consider the system \eqref{equ:1DCase:OriginalSystem}. Suppose $f(x;\mu_{1},\mu_{2})\in C^{2}(\mathbb{R}\times \mathbb{R}^{2},\mathbb{R})$ with a nonhyperbolic equilibrium at the origin, $f(0;0,0)=0$, $D_{x}f(0;0,0)=0$, and satisfying the nondegeneracy condition
    \begin{equation*}
        D_{xx}f(0;0,0)\neq 0,
    \end{equation*}
    and transversality condition for either $\mu_{1}$ or $\mu_{2}$
    \begin{equation*}
        D_{\mu_{1}}f(0;0,0)\neq 0 \quad \mathrm{or}\quad D_{\mu_{2}}f(0;0,0)\neq 0,
    \end{equation*}
    i.e., the system \eqref{equ:1DCase:OriginalSystem} has a saddle-node bifurcation where either $\mu_{1}$ or $\mu_{2}$ is the bifurcation parameter. This defines a one-dimensional smooth curve $\Gamma\subset \mathbb{R}^{3}$ in a neighbourhood of $(x,\mu_{1},\mu_{2})=(0,0,0)$ in which $f$ satisfies the singularity and nondegeneracy conditions.

    Consider the extended system \eqref{equ:1DCase:ExtendedSystem} by transforming parameter $\mu_{1}$ into a variable, where $f\in C^{2}(\mathbb{R}^{2}\times \mathbb{R},\mathbb{R})$ and $g\in C^{2}(\mathbb{R}\times\mathbb{R}, \mathbb{R})$. If there is a point $(x,\mu_{1},\mu_{2})=(x^{*},\mu_{1}^{*},\mu_{2}^{*})\in \Gamma$ such that $g(\mu_{1};\mu_{2})$ satisfies the singularity conditions
    \begin{equation}
        g(\mu_{1}^{*}; \mu_{2}^{*})=0, \quad D_{\mu_{1}}g(\mu_{1}^{*};\mu_{2}^{*})=b\neq 0,
        \label{equ:1DCase:Proposition1:SingularityConditions}
    \end{equation}
    and the transversality condition
    \begin{equation}
        \det\begin{pmatrix}
            D_{\mu_{1}}f & D_{\mu_{1}}g \\
            D_{\mu_{2}}f & D_{\mu_{2}}g
        \end{pmatrix} = D_{\mu_{1}}fD_{\mu_{2}}g - D_{\mu_{1}}gD_{\mu_{2}}f \neq 0,
      \label{equ:1DCase:Proposition1:TransversalityCondition}
    \end{equation}
    at $(x,\mu_{1},\mu_{2})=(x^{*},\mu_{1}^{*},\mu_{2}^{*})$, then the extended system  \eqref{equ:1DCase:ExtendedSystem} has a saddle-node bifurcation at $(x,\mu_{1})=(x^{*},\mu_{1}^{*})$ as $\mu_{2}$ crosses $\mu_{2}^{*}$.

    Moreover, there exists a unique function $\mu_{1}=\nu(\mu_{2})$ such that $\mu_{1}^{*}=\nu(\mu_{2}^{*})$, and the extended system is reduced to one dimension around $(x^{*},\mu_{1}^{*})$
    \begin{equation*}
        \dot \xi=f(\xi+\tfrac{a}{b}(\nu(\mu_{2})-\mu_{1}^{*})+x^{*}, \nu(\mu_{2});\mu_{2})-\tfrac{a}{b}g(\nu(\mu_{2});\mu_{2}),
    \end{equation*}
    where $\xi=x-x^{*}-\frac{a}{b}(\mu_{1}-\mu_{1}^{*})$ and $a=D_{\mu_{1}}f(x^{*},\mu_{1}^{*};\mu_{2}^{*})$.
\end{proposition}
\begin{proof}
    Let $z=(x, \mu_{1})^{T}$ and $F(z;\mu_{2})=F(x,\mu_{1};\mu_{2})=(f(x,\mu_{1};\mu_{2}),g(\mu_{1};\mu_{2}))^{T}$. By definition of $\Gamma$, the singularity conditions ($f=0$ and $D_{x}f=0$), the nondegeneracy condition ($D_{xx}f\neq0$), and one of the transversality conditions ($D_{\mu_{1}}f\neq0$ or $D_{\mu_{2}}f\neq0$) are satisfied at the point $(z^{*},\mu_{2}^{*})=(x^{*},\mu_{1}^{*},\mu_{2}^{*})\in\Gamma$. Since $g(\mu_{1}^{*};\mu_{2}^{*})=0$, we also have $F(x^{*},\mu_{1}^{*};\mu_{2}^{*})=0$ (first singularity condition for $F$). The Jacobian of $F$ evaluated at $z^{*}$ is
    \begin{equation*}
    A=D_{z}F(x^{*},\mu_{1}^{*};\mu_{2}^{*})=
    \left.\begin{pmatrix}
        D_{x}f & D_{\mu_{1}}f \\
        D_{x}g & D_{\mu_{1}}g
    \end{pmatrix}\right|_{z=z^{*}}=
    \begin{pmatrix}
        0 & a \\
        0 & b
    \end{pmatrix},
    \end{equation*}
    where $a=D_{\mu_{1}}f(x^{*},\mu_{1}^{*};\mu_{2}^{*})$ and $b=D_{\mu_{1}}g(\mu_{1}^{*};\mu_{2}^{*})\neq0$, by assumption \eqref{equ:1DCase:Proposition1:SingularityConditions}. Since $\det(A)=0$ and $\mathop\mathrm{tr}(A)=b\neq 0$, the eigenvalues of $A$ are $\lambda_{1}=0$ and $\lambda_{2}=b\neq 0$ with corresponding eigenvectors
    \[v_{\lambda_{1}}=\begin{pmatrix} 1 \\ 0 \end{pmatrix},
    \quad v_{\lambda_{2}}=\begin{pmatrix} a \\ b \end{pmatrix}.\]
    Note that condition $D_{\mu_{1}}g\neq0$ is needed to guarantee only one zero eigenvalue. Thus, $D_{z}F$ is singular with only one zero eigenvalue (second singularity condition for $F$).

    The diagonalization matrix $P$ and its inverse are given by
    \[P=\begin{pmatrix}
         1 & a \\
         0 & b
        \end{pmatrix}, \quad
    P^{-1}=\frac{1}{b}\begin{pmatrix}
         b & -a \\
         0 & 1
        \end{pmatrix}.
    \]
    Let the new shifted coordinates be defined by
    \begin{equation*}
        \begin{pmatrix}
            \xi \\
            \upsilon
        \end{pmatrix} = P^{-1}
        \begin{pmatrix}
            x-x^{*} \\
            \mu_{1} - \mu_{1}^{*}
        \end{pmatrix} =
        \begin{pmatrix}
            x -x^{*} - \frac{a}{b}(\mu_{1}-\mu_{1}^{*}) \\
            \frac{1}{b}(\mu_{1}-\mu_{1}^{*})
        \end{pmatrix}.
    \end{equation*}
    Then the corresponding extended system is given by
    \begin{equation*}
        \begin{aligned}
            \dot \xi &= f(\xi+a\upsilon+ x^{*}, b\upsilon + \mu_{1}^{*};\mu_{2}) - \tfrac{a}{b}g(b\upsilon + \mu_{1}^{*};\mu_{2}^{*}), \\
            \dot \upsilon & = \tfrac{1}{b}g(b\upsilon+\mu_{1}^{*};\mu_{2}^{*}).
        \end{aligned}
    \end{equation*}
    Define
    \begin{equation}
        f_{1}(\xi,\upsilon;\mu_{2})=f(\xi+a\upsilon+ x^{*}, b\upsilon + \mu_{1}^{*};\mu_{2}) - \tfrac{a}{b}g(b\upsilon + \mu_{1}^{*};\mu_{2}^{*}),
        \label{equ:1DCase:Proposition1:Proof:f1}
    \end{equation}
    and define $f_{2}(\upsilon; \mu_{2})$ such that
    \begin{equation*}
        \tfrac{1}{b}g(b\upsilon+\mu_{1}^{*};\mu_{2}^{*}) = b \upsilon + f_{2}(\upsilon;\mu_{2}).
    \end{equation*}
    Then,
    \begin{equation}
        \begin{aligned}
            \dot \xi &= f_{1}(\xi ,\upsilon;\mu_{2}), \\
            \dot \upsilon &= b \upsilon + f_{2}(\upsilon; \mu_{2}).
        \end{aligned}
        \label{equ:1DCase:Proposition1:Proof:Coordinates}
    \end{equation}

    Note that the singularity conditions are satisfied by construction,
    \begin{equation}
        \begin{aligned}
        & f_{1}(0,0;\mu_{2}^{*})=f_{2}(0;\mu_{2}^{*})=0, \\
%         & D_{\zeta}f_{1}(0,0;\mu_{2}^{*})=D_{\zeta}f_{2}(0,0;\mu_{2}^{*})=0, \\
        & D_{\xi}f_{1}(0,0;\mu_{2}^{*})=D_{\upsilon}f_{1}(0,0;\mu_{2}^{*})=D_{\xi}f_{2}(0;\mu_{2}^{*})=D_{\upsilon}f_{2}(0;\mu_{2}^{*})=0.
%         & D_{\xi}f_{1}(0,0;\mu_{2}^{*})=D_{x}f(x^{*},\mu_{1}^{*},\mu_{2}^{*})=0 \\
%         & D_{\nu}f_{1}(0,0;\mu_{2}^{*})=D_{x}f(x^{*},\mu_{1}^{*},\mu_{2}^{*}) a + D_{\mu_{1}}f(x^{*},\mu_{1}^{*},\mu_{2}^{*})b-\tfrac{a}{b}D_{\mu_{1}}gb=0
        \end{aligned}
        \label{equ:1DCase:Proposition1:Proof:NoLinearTerms}
    \end{equation}
    The nondegeneracy condition for $f_{1}$ is satisfied since
    \begin{equation*}
        D_{\xi\xi} f_{1}(0,0;\mu_{2}^{*}) = D_{xx}f(x^{*},\mu_{1}^{*};\mu_{2}^{*}))\neq0.
    \end{equation*}
    The transversality condition for $f_{1}$ follows from dividing the determinant in~\eqref{equ:1DCase:Proposition1:TransversalityCondition} by $-b\neq0$ and the definition of $f_{1}$ \eqref{equ:1DCase:Proposition1:Proof:f1}
    \begin{gather*}
        ( \cancelto{a}{D_{\mu_{1}}f(x^{*},\mu_{1}^{*};\mu_{2}^{*})} D_{\mu_{2}}g(\mu_{1}^{*};\mu_{2}^{*}) - \cancelto{b}{D_{\mu_{1}}g(\mu_{1}^{*};\mu_{2})} D_{\mu_{2}}f(x^{*},\mu_{1}^{*};\mu_{2}^{*}))\neq0,\\
        \implies D_{\mu_{2}}f(x^{*},\mu_{1}^{*};\mu_{2}^{*}) - \tfrac{a}{b}D_{\mu_{2}}g(\mu_{1}^{*};\mu_{2}^{*}) = D_{\mu_{2}}f_{1}(0,0;\mu_{2}^{*}) \neq 0.
    \end{gather*}
    Then, by Theorem~\ref{the:SaddleNodeBifurcationND} and Corollary~\ref{the:SaddleNodeBifurcationNDCorollary}, the transformed system \eqref{equ:1DCase:Proposition1:Proof:Coordinates} has a saddle-node bifurcation point at $(0,0)$ as $\mu_{2}$ crosses $\mu_{2}^{*}$.

    Transforming back to the variable $z$, we have that the extended system \eqref{equ:1DCase:ExtendedSystem} has a saddle-node bifurcation point at $(x^{*},\mu_{1}^{*})$ as $\mu_{2}$ crosses $\mu_{2}^{*}$.

    Now, denote
    \begin{equation*}
        F_{2}(\upsilon;\mu_{2})=b\upsilon +f_{2}(\upsilon;\mu_{2})=0.
    \end{equation*}
    Note that $D_{v}F_{2}(0;\mu_{2}^{*})=b\neq 0$, by equation \eqref{equ:1DCase:Proposition1:Proof:NoLinearTerms}. By the Implicit Function Theorem~\ref{the:ImplicitFunction}, there is a neighbourhood of $\mu_{2}=\mu_{2}^{*}$ where there exists a unique function $\upsilon=\hat\nu(\mu_{2})$ such that $\hat\nu(\mu_{2}^{*})=0$ and $F_{2}(\eta_{}(\mu_{2}),\mu_{2})=0$. Then, equation \eqref{equ:1DCase:Proposition1:Proof:Coordinates} reduces to
    \begin{equation*}
        \dot \xi = f_{1}(\xi ,\hat\nu(\mu_{2});\mu_{2}).
    \end{equation*}
    Changing back to $\mu_{1}$, we have
    \begin{gather*}
        \upsilon = \tfrac{1}{b}(\mu_{1}-\mu_{1}^{*}) \\
        \implies \mu_{1}=b\upsilon+\mu_{1}^{*}=b\hat\nu(\mu_{2})+\mu_{1}^{*}.
    \end{gather*}
    Define $\mu_{1}=\nu(\mu_{2}) = b\hat\nu(\mu_{2})+\mu_{1}^{*}$, then $\nu(\mu_{2}^{*})=b\hat\nu(\mu_{2}^{*})+\mu_{1}^{*}=\mu_{1}^{*}$.
    Finally, using the definition of $f_{1}(\xi,\nu,\mu_{2})$, we have
    \begin{align*}
        \dot \xi & = f_{1}(\xi ,\nu(\mu_{2});\mu_{2}) \\
        & = f_{1}(\xi,\tfrac{1}{b}(\nu(\mu_{2})-\mu_{1}^{*});\mu_{2}) \\
        & = f(\xi+\tfrac{a}{b}(\nu(\mu_{2})-\mu_{1}^{*})+x^{*}, \nu(\mu_{2});\mu_{2})-\tfrac{a}{b}g(\nu(\mu_{2});\mu_{2}) \\
        & = f(\xi+\tfrac{a}{b}(\nu(\mu_{2})-\mu_{1}^{*})+x^{*}, \nu(\mu_{2});\mu_{2})-\tfrac{a}{b}g(\nu(\mu_{2});\mu_{2}).
    \end{align*}
\end{proof}

This theorem provides a way to extend the scalar system \eqref{equ:1DCase:OriginalSystem} where a saddle-node occurs by transforming any parameter, $\mu_{1}$ for convenience, into a variable to obtain the extended system \eqref{equ:ExtendedSystem} where a saddle-node bifurcation now occurs as the other parameter, $\mu_{2}$, passes through some bifurcation value $\mu_{2}^{*}$. Note that the transformed parameter, $\mu_{1}$, does not need to be the original bifurcation parameter. Thus, we say that the saddle-node bifurcation in the extended system is the carryover of the saddle-node bifurcation in the original system. Also note that Proposition~\ref{the:1DCase:SaddleNodeBifurcationExtension1D} requires that $g(\mu_{1};\mu_{2})$ does not depend explicitly on $x$. This makes the conditions of this proposition easy to verify with graphical and numerical tools.

\begin{proposition}
    \label{the:1DCase:SaddleNodeBifurcationExtension1DCorollary}
    Under the conditions of Proposition~\eqref{the:1DCase:SaddleNodeBifurcationExtension1D}, let $h(\mu_{1},\mu_{2})=0$ be the projection of $\Gamma$ onto the $(\mu_{1},\mu_{2})$-plane. If $h(\mu_{1},\mu_{2})$ is differentiable at $(\mu_{1}^{*},\mu_{2}^{*})$, then conditions \eqref{equ:1DCase:Proposition1:TransversalityCondition} and \eqref{equ:1DCase:Proposition1:SingularityConditions} are equivalent to
    \begin{enumerate}
        \item $g(\mu_{1};\mu_{2})=0$ intersects $h(\mu_{1},\mu_{2})=0$ transversally at a point $(\mu_{1}^{*},\mu_{2}^{*})$, and
        \item the tangent line to $g(\mu_{1};\mu_{2})=0$ at $(\mu_{1}^{*},\mu_{2}^{*})$ is not parallel to the $\mu_{1}$-axis,
    \end{enumerate}
    respectively.
\end{proposition}

This proposition says that in order to find the saddle-node bifurcation points for the extended system, we plot the two-parameter bifurcation diagram of the smaller system, superimpose the nullclines of the new equation in the extended system, and look for transverse intersections between the saddle-node bifurcation curve and the nullclines. This is enough to verify the singularity and transversality conditions in the extended system.

\begin{proof}[Proof of Proposition~\ref{the:1DCase:SaddleNodeBifurcationExtension1DCorollary}]
    Note that $(\mu_{1}^{*},\mu_{2}^{*})$ satisfies $g=0$. Now, two vectors $u,v\in\mathbb{R}^{2}$ are transverse (parallel) if and only if the determinant of the matrix formed by them is non-zero (is zero), i.e.,
    \[det(u,v)=u_{1}v_{2}-v_{1}u_{2}=|u||v|\sin(\theta)\neq 0 \iff \theta \neq 0,\pi.\]

    Recall that $h(\mu_{1},\mu_{2})=0$ is defined by the projection of $\Gamma$ onto the $(\mu_{1},\mu_{2})$-plane, given by $(x,\mu_{1},\mu_{2},)\mapsto(\mu_{1},\mu_{2})$. Since at least one of $D_{\mu_{1}}f$ or $D_{\mu_{2}}f$ is non-zero, points on $\Gamma$ have a unique correspondence to points on $h(\mu_{1}, \mu_{2})=0$. Thus, a point $(\mu_{1}^{*},\mu_{2}^{*})$ at which $g=0$ and $h=0$ intersect has a unique corresponding point $(z^{*},\mu_{2}^{*})=(x^{*},\mu_{1}^{*}, \mu_{2}^{*})\in\Gamma$.

    Assume $D_{\mu_{1}}f\neq0$ at $(z^{*};\mu_{2}^{*})$. Then, by the Implicit Function Theorem~\ref{the:ImplicitFunction}, we can parameterize $\Gamma$ by $\mu_{2}$ with functions $x=\mathcal{X}(\mu_{2})$ and $\mu_{1}=\mathcal{M}(\mu_{2})$ such that $x^{*}=\mathcal{X}(\mu_{2}^{*})$ and $\mu_{1}^{*}=\mathcal{M}(\mu_{2}^{*})$ (see equation~\eqref{equ:1DCase:XandMfunctions}). Implicit differentiation of $f(x,\mu_{1};\mu_{2})=0$ with respect to $\mu_{2}$ gives
    \[D_{x}f\mathcal{X}'+D_{\mu_{1}}f\mathcal{M}'+D_{\mu_{2}}f=0.\]
    At $z^{*}$, $D_{x}f=0$ and we have
    \[\mathcal{M}'=-\frac{D_{\mu_{2}}f}{D_{\mu_{1}}f}.\]
    Implicit differentiation of $h(\mu_{1},\mu_{2})=0$ with respect to $\mu_{2}$ gives 
    \[D_{\mu_{1}}h\mathcal{M}'+D_{\mu_{2}}h=0.\]
    Evaluating at $z^{*}$, substituting the $\mathcal{M}'$ and multiplying by $-D_{\mu_{1}}f$, we obtain
    \begin{equation*}
        D_{\mu_{1}}hD_{\mu_{2}}f-D_{\mu_{2}}hD_{\mu_{1}}f=0.
    \end{equation*}
    This means that vectors $(D_{\mu_{1}}f,D_{\mu_{2}}f)^{T}$ and $(D_{\mu_{1}}h,D_{\mu_{2}}h)^{T}$ are multiple of each other at $z^{*}$. Note that this is also true if $D_{\mu_{1}}f=0$ since we must have $D_{\mu_{2}}f\neq0$ and similar arguments follow. Thus, $(D_{\mu_{1}}h,D_{\mu_{2}}h)^{T}$ and $(D_{\mu_{1}}g,D_{\mu_{2}}g)^{T}$ are transverse if and only if $(D_{\mu_{1}}f,D_{\mu_{2}}f)^{T}$ and $(D_{\mu_{1}}g,D_{\mu_{2}}g)^{T}$ are transverse, which is equivalent to saying that the transversality condition \eqref{equ:1DCase:Proposition1:TransversalityCondition} holds.

>>>>

\title{Carryover of a saddle-node bifurcation after transformation of a parameter into a variable: Appendix notes}
\author{Carlos Contreras, Gustavo Carrero, Gerda de Vries}
% \address{University of Alberta}

@@

Detailed calculations of examples in the main paper.

\section{Gen activation model}

@@
